Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-08-14 paper/main.tex

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

@@

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how EDSLs may be constructed to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

@@

%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it: this would, however, cause
problems with later steps.

@@

We structure the remainder of this paper as follows. We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.

@@

\section{An overview of the \texttt{Effects} framework}
\label{effects}
\texttt{Effects} is an EDSL which handles side-effects such as database access
as \textit{algebraic effects}, allowing greater code modularity. This section
is intended to serve as a brief introduction to the framework; for more
information, see \cite{brady:effects}.

An effect is represented as an instance of the \texttt{EFFECT} type. Such
instances may be associated with a \textit{resource}, Describing a program in
the \texttt{Effects} library is achieved using the following data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}
\texttt{EffM} is a type that is parameterised over an execution context
(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
of effects at the end of the program's execution (\texttt{es'}), and a return
type (\texttt{a}). The list of input types can be used to ensure that an
operation may only be called when the program is in a particular state, and
since the list of input effects may differ from the list of output effects, we
may modify the resource associated with the state after each operation. This
allows us to simply enforce a resource usage protocol, and deviation from this
protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind
\texttt{Effect}, which describes the available operations and how they modify
the resource associated with the effect. Each operation consists of an initial
resource \texttt{res}, the resource at the end of the operation \texttt{res'},
and the output type of the operation \texttt{t}.

%TODO: Execution
\begin{Verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{Verbatim}

Operations are made available to users by writing functions with the
corresponding \texttt{EffM} type, which may then be used to invoke the
operation from within an effectful program.

\begin{Verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{Verbatim}

Additionally, we define \textit{handlers}, which describe how each operation is
interpreted in a given execution context. Finally, \texttt{Effect} instances
are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
constructor, which is provided with the initial resource value.

@@

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses. Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.

@@

\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

@@

Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library. This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.

\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

@@

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels
\begin{Verbatim}[samepage]
data Session : Effect where
  LoadSession : 
  SessionID -> 
  Session (SessionRes SessionUninitialised) 
          (SessionRes SessionInitialised) 
          (Maybe SessionData)
  UpdateSession : 
    SessionData -> 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) ()
  CreateSession : 
    SessionData -> 
    Session (SessionRes SessionUninitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  DeleteSession : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool 
  WriteToDB : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool
  DiscardSessionChanges : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) ()
  GetSessionID : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  GetSessionData : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionData)
\end{Verbatim}
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}
Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler. In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key. The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage. This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
\begin{Verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

@@

This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker. One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data. This may involve sending an integer detailing the current page, which could be used in a query such as:

@@

In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information. Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.

@@

In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
\subsection{Form Construction}
\label{formcons}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.

\idris{} allows for implicit arguments to be bound across a block of code through \texttt{using} notation. We may therefore parameterise the EDSL over the types associated with each form element, and the effects required by the handler function.

@@

    Submit : (mkHandlerFn ((reverse G), E)) ->
             String -> 
             Form (FormRes G E) (FormRes [] []) String
\end{Verbatim}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.

We make use of the resource associated with the EDSL, \texttt{FormRes}, to construct the form. The resource allows us to record the types associated with each form element and the HTML required to display the form. he constructor, \texttt{FR}, requires the number of elements in the form in order to allow for the naming of new elements, the list of element types, the list of effects supported by the handler function, and the currently generated HTML for the form.

@@

To specify a form instance, we define a function of type \texttt{UserForm}:
\begin{Verbatim}
  UserForm : Type
  UserForm = EffM m [FORM (FormRes []) 
                          (FormRes [])] String
\end{Verbatim}
All forms are required to include a submit button, as mandated by the requirement that the input and output resource contains an empty list of types; this requirement is fulfilled as per the output resource type of the \texttt{AddSubmit} operation. As the creation of a form is a pure function which does not include side effects, we do not restrict the handler to IO as with previously-discussed EDSLs, instead denoting the fact that it may be run in any handler with the implicit variable \texttt{m}.

@@

Running the form construction EDSL is achieved as an operation of the CGI effect, \texttt{AddForm}, which then outputs the generated HTML to the page. The generated metadata describing the handler function is serialised as a hidden HTML field.

@@

WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres. WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes. The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables. These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy. Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation. We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.

>>>>

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

@@

In this paper, we apply this technique to web programming, and introduce an
EDSL, to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

@@

%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it. However, this would cause
problems with later steps.

@@

We structure the remainder of this paper as follows. We provide a brief
overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
how this may be used to ensure adherence to resource usage protocols for CGI,
SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
type-safe form handling in Section ~\ref{form}, implemented using
\texttt{Effects}; and discuss the larger example
of a message board system making use of these components in Section
~\ref{messageboard}.

@@

\section{An overview of the \texttt{Effects} framework}
\label{effects}

\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
side-effects such as state, exceptions, and I/O as \emph{algebraic
effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
by an input and output \emph{state}, which permits effectful programs to track
the progress of a resource usage protocol. Effectful programs are written
in a monadic style, with \texttt{do}-notation, with their type stating which
specific effects are available.
Effectful
programs are described using the following data type,
in the simplest case:

\begin{Verbatim}
Eff : (m : Type -> Type) ->
      (es : List EFFECT) -> (a : Type) -> Type
\end{Verbatim}

\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
which describes the context in which the effectful program will be run, a
list of side effects \texttt{es} that the program is permitted to use, and the
program’s return type \texttt{a}. The name \texttt{m} for the computation context is
suggestive of a monad, but there is no requirement for it to be so.

For example, the following type carries an integer state,
throws an exception of type \texttt{String} if the state reaches 100, 
and runs in a \texttt{Maybe} context:

\begin{Verbatim}
addState : Eff Maybe [STATE Int, EXCEPTION String] ()
addState = do val <- get
              when (val == 100) (raise "State too big")
              put (val + 1)
\end{Verbatim}

\subsection{Implementing Effects}

Effects such as state and exception are described as algebraic data types,
and run by giving \emph{handlers} for specific computation contexts. 
Effects have a corresponding \emph{resource} (in the case of state, the
resource is simply the current state). Executing an effectful operation may
change the resource and return a value:

\begin{Verbatim}
Effect : Type
Effect = (in_res : Type) -> (out_res : Type) -> 
         (val : Type) -> Type
\end{Verbatim}

For example, the state effect is described as follows:

\begin{Verbatim}
data State : Effect where
     Get :      State a a a
     Put : b -> State a b ()
\end{Verbatim}

That is, \texttt{Get} returns a value of type \texttt{a} without updating
the resource type. \texttt{Put} returns nothing, but has the effect of updating
the resource. In order to make an effect usable, we implement a handler
for a computation context by making an instance of the following class:

\begin{Verbatim}
class Handler (e : Effect) (m : Type -> Type) where
     handle : res -> 
              (eff : e res res' t) -> 
              (k: res' -> t -> m a) -> m a
\end{Verbatim}

The \texttt{handle} function takes the input resource, an effect which may
update that resource and execute a side-effect, and a continuation \texttt{k}
which takes the updated resource and the return value of the effect. We use
a continuation here primarily because there is no restriction on the number of
times a handler may invoke the continuation (raising an exception, for example,
will not call the continuation). Reading and updating states in handled
for all computation contexts \texttt{m} as follows:

\begin{Verbatim}
instance Handler State m where
     handle st Get     k = k st st
     handle st (Put n) k = k n ()
\end{Verbatim}

Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
the \texttt{Get} and \texttt{Put} operations into functions in \texttt{Eff}, as
follows:

\begin{Verbatim}
STATE : Type -> EFFECT
STATE t = MkEff t State

get : Eff m [STATE x] x
get = Get

put : x -> Eff m [STATE x] ()
put val = Put val
\end{Verbatim}

A concrete effect is simply an algebraic effect type paired with its current
resource type. This, and other
technical details, are explained in full elsewhere~\cite{brady:effects}.
For the purposes of this paper, it suffices to know how to describe and
handle new algebraic effects.

\subsection{Resource Protocols as Effects}

More generally, a program might modify the set of effects available.
This might be desirable for several reasons, such as adding a new
effect, or to update an index of a dependently typed state. In this
case, we describe programs using the \texttt{EffM} data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) -> (es' : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}

\texttt{EffM} is parameterised over the context and type as before, but
separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
In fact, \texttt{Eff}
is defined in terms of \texttt{EffM}, with equal input/output effects.
We can use this to describe how effects follow resource protocols. A simple
example is a file access protocol, where a file must be opened before it
is read or written, and a file must be closed on exit:

\begin{Verbatim}
data FileIO : Effect where
     Open  : String -> (m : Mode) -> 
             FileIO () (OpenFile m) ()
     Close : FileIO (OpenFile m) () ()

     ReadLine  : FileIO (OpenFile Read)  
                        (OpenFile Read) String
     WriteLine : String -> 
                 FileIO (OpenFile Write) 
                        (OpenFile Write) ()
     EOF       : FileIO (OpenFile Read)  
                        (OpenFile Read) Bool
\end{Verbatim}

Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file changes an empty resource to
a resource containing an open file; reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions, as well as throwing an exception
if any operation fails.

The following program type checks, and therefore implicitly carries 
a proof that the file resource protocol is followed correctly:

\begin{Verbatim}
testFile : Eff IO [FILE_IO (), STDIO] () 
testFile = catch (do open "testFile" Read
                     str <- readLine
                     close
                     putStrLn str)
                 (\err => putStrLn ("Error: " ++ show err))
\end{Verbatim}

The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit.
Therefore, attempting to write to the file, or failing to open or close the
file, would cause a \emph{compile-time} error.

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effet, implicitly carry a proof that the resource
protocol has been correctly followed.

@@

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource.

@@

\begin{figure}[h]
\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
\caption{CGI Effect}
\label{fig:cgieffect}
\end{figure}

Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. This is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity.

\begin{figure}[h]
\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
\caption{Database Effect}
\label{fig:dbeffect}
\end{figure}

There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

@@

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels

\begin{figure}[h]
\begin{Verbatim}[samepage]
data Session : Effect where
  LoadSession : 
  SessionID -> 
  Session (SessionRes SessionUninitialised) 
          (SessionRes SessionInitialised) 
          (Maybe SessionData)
  UpdateSession : 
    SessionData -> 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) ()
  CreateSession : 
    SessionData -> 
    Session (SessionRes SessionUninitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  DeleteSession : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool 
  WriteToDB : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool
  DiscardSessionChanges : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) ()
  GetSessionID : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  GetSessionData : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionData)
\end{Verbatim}
\caption{Session Effect}
\label{fig:sessioneffect}
\end{figure}

\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
algebraic effect. In this application, there exist two states:
In \texttt{SessionUninitialised}, the user may load or create a
session.
In \texttt{SessionInitialised}, the user may update the
representation of the session in memory, serialise the session and write it to
the database, or delete the session and invalidate the user's session key. The
introduction of these two states ensures that changes are explicitly either
written or discarded, eliminating the possibility of a developer updating the
session but neglecting to commit it to persistent storage. This, of course, is
under the assumption that the process exits cleanly: we attempt to facilitate
this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the
resource associated with the effect.

\begin{Verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

@@

This can in turn lead to risks; developers may assume that data is
of a certain type, and therefore discount the possibility that it may have been
modified by an attacker. One example would be the traversal of paginated data,
in which a form is used to make a request to retrieve the next page of data.
This may involve sending an integer detailing the current page, which could be
used in a query such as:

@@

In this section, we present a mechanism by which we introduce a DSL
for the creation of web forms which preserve type information, implemented
as a dependent algebraic effect. Once the form has
been submitted, retrieved information is passed directly to a
developer-specified function for handling, without the need to manually check
and deserialise data.

@@

In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
of the form-handling system: namely, the effect which allows the creation of
forms, and the handling code which deserialises the data and passes it to the
user-specified handler function.

\subsection{Form Construction}
\label{formcons}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: the specification of an
incompatible handler will result in a compile-time type error.

\idris{} allows for implicit arguments to be bound across a block of code
through \texttt{using} notation. We may therefore parameterise the 
form data types over
the types associated with each form element, and the effects required by the
handler function.

@@

    Submit : (mkHandlerFn ((reverse G), E)) ->
             String -> 
             Form (FormRes G E) (FormRes [] []) String
\end{Verbatim}
The implementation of the form effect also contains other constructs for
additional form elements such as radio buttons and check boxes, but are omitted
in the interest of brevity.

We make use of the resource associated with the effect, \texttt{FormRes}, to
construct the form. The resource allows us to record the types associated with
each form element and the HTML required to display the form. he constructor,
\texttt{FR}, requires the number of elements in the form in order to allow for
the naming of new elements, the list of element types, the list of effects
supported by the handler function, and the currently generated HTML for the
form.

@@

To specify a form instance, we define a function of type \texttt{UserForm}:
\begin{Verbatim}
  UserForm : Type
  UserForm = EffM m [FORM (FormRes []) 
                          (FormRes [])] String
\end{Verbatim}
All forms are required to include a submit button, as mandated by the
requirement that the input and output resource contains an empty list of types;
this requirement is fulfilled as per the output resource type of the
\texttt{AddSubmit} operation. As the creation of a form is a pure function
which does not include side effects, we do not restrict the handler to IO as
with previously-discussed effects, instead denoting the fact that it may be run
in any handler with the implicit variable \texttt{m}.

@@

Running form construction is achieved as an operation of the CGI
effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
The generated metadata describing the handler function is serialised as a
hidden HTML field.

@@

WebDSL \cite{webdsl} is a domain-specific language written primarily to
introduce new abstractions which aim to reduce the amount of boilerplate code
that must be written and maintained by developeres. WebDSL is built on top of
Java, which often includes a large amount of redundant code such as accessor
and mutator functions within entity classes. The Java Persistence API (JPA)
\cite{jpa} provides an object-relational mapping through the use of Java 5
annotations, which may then be used to construct database tables. These
annotations soon become complex, however, and coupled with redundant
boilerplate code, data model declarations may soon become unwieldy. Through the
use of WebDSL, these data model declarations can be much more compactly
declared, and elaborated into Java code by parsing the data-modelling DSL into
an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL
also applies similar concepts to implement a \textit{template system} for such
objects, which allows the data to be used in code generation. We look to
implement many of these ideas, but as effects within the IdrisWeb
framework, as with the form construction effect.
